Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-05-13 manuscript/cpcrw_manuscript.Rmd
Many minor tweaks to intro, methods, results

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Carbon dioxide fluxes were strongly influenced by incubation chamber temperature, core water content, and percent soil nitrogen. Methane fluxes were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~. The CO~2~ temperature sensitivity (i.e. Q~10~) was `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively. These results suggest that deep active-layer soils may be much more sensitive to changes in moisture than to temperature, a critical factor as discontinuous permafrost melts in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results help us understand the future dynamics, and feedback potential, of such soils.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented increased fluxes of CO~2~, and under some conditions of CH~4~ {Olefeldt, 2013 #3456}, with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. Greenhouse gas (GHG) responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} may counteract these effects. A critical question, then, is how the structure, chemistry, and microbial communities of current active-layer soils will respond to almost-certain temperature and moisture changes in the future {Xue, 2016 #4361}.

@@

The field component of this research took place in Caribou-Poker Creeks Research Watershed (CPCRW), part of the Bonanza Creek LTER (http://www.lter.uaf.edu/research/study-sites-cpcrw). CPCRW is located in the Yukon-Tanana Uplands northeast of Fairbanks, AK, a part of the boreal forest that has seen strong increases in air temperature and forest browning {Ju, 2016 #4320} over several decades. Annual average air temperature was -2.5 °C, and annual average precipitation 400 mm {Petrone, 2006 #4313}. The watershed's lowlands and north-facing slopes are dominated by black spruce (*Picea mariana* (Mill.) BSP), feathermoss (*Pleurozium schreberi* and others), and *Sphagnum* spp.; the drier south slopes tend to be deciduous with a mixture of trembling aspen (*Populus tremuloides* Michx.), paper birch (*Betula neoalaskana*), and patches of alder (*Alnus crispa*).

@@

Thirty-nine soil cores, each 30 cm high by 7.5 cm wide, were taken using a soil recovery augur (AMS Inc., American Falls, ID) on 3-5 August 2015. We sampled from the bottom (within 0-2 cm) of the active layer, which averaged 80 cm depth. Sample points were randomly located in the transition zone described above, and separated by 2-5 m. Cores were kept cool in the field before being packed in dry ice and shipped to the lab in Richland, WA within 48-72 hours of collection.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron Control Systems BDW80, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) emissions four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores with a 200 µm mesh screen fit to the base were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. When necessary, cores received additional wetting from the top to maintain their water status at the desired level.

@@

At each GHG measurement, we measured the rise in gas concentrations, considering this as the flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~). The GHG concentration measurements for each sample (i.e., each gas, core, and date/time) were used to calculate a linear rate of change (δc/δt), based on the concentration rise from a minimum (up to `r MAX_MINCONC_TIME` seconds after measurement began) to a maximum (at `r MAX_MINCONC_TIME`-`r MAX_MAXCONC_TIME` seconds). This value was calculated for each sample, i.e. each individual core measurement throughout the 100-day incubation.

@@

Each core’s respiration flux (*F*) was calculated following e.g. Steduto et al. {, 2002 #3383} as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a dry soil mass basis (µg C g soil^-1^ day^-1^).

@@

Cumulative respiration for each core and gas was calculated by linearly interpolating flux rates between measurement dates and summing respired C over the entire incubation. The effect of temperature and treatment (drought, controlled drought, or field moisture conditions) on cumulative gas fluxes was evaluated with a post-hoc Tukey Honest Significant Differences test. Temperature sensitivity (Q~10~) was calculated for each gas and treatment as $\frac{F_2}{F_1}^{\left( \frac{10}{T_2-T_1} \right)}$ where *F~1~* and *F~2~* are the cumulative gas fluxes (mg C) at temperatures *T~1~* and *T~2~* (°C), respectively.

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI to be filled in on acceptance].

@@

Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (min `r pn(gwcbegin$min, 2)`, max `r pn(gwcbegin$max, 2)`) at the beginning of the incubation (**Figure 1**). "Field moisture" cores were on average unchanged (`r pn(gwcendmean["Field moisture"], 2)` ± `r pn(gwcendsd["Field moisture"], 2)`) at the end of the incubation, but both the drought treatments, which did not differ from each other in their effect on gravimetric water content (P = `r pn(p_trt_wc, 3)`), had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Volumetric water content values ranged from `r pn(vwcbegin$mean, 2)` ± `r pn(vwcbegin$sd, 2)` (min `r pn(vwcbegin$min, 2)`, max `r pn(vwcbegin$max, 2)`) at the beginning of the experiment to `r pn(vwcend$mean, 2)` ± `r pn(vwcend$sd, 2)` (min `r pn(vwcend$min, 2)`, max `r pn(vwcend$max, 2)`) at the end across all cores. Water filled pore space, assuming a particle density of `r PARTICLE_DENSITY` g cm^-3^, was `r pn(min(fluxdata$WFPS_percent), 0)`-`r pn(max(fluxdata$WFPS_percent), 0)`%.

@@

Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

@@

We found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were highly sensitive to temperature and drought effects. Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664}. Waldrop et al. {, 2010 #4380} observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.

>>>>

Rapid climatic changes, rising air temperatures, and increased fires are expected to drive permafrost degradation and alter soil carbon (C) cycling in many high-latitude ecosystems. How these soils will respond to changes in their temperature, moisture, and overlying vegetation is highly uncertain, but critical to understand given the large soil C stocks in these regions. We used a laboratory experiment to examine how temperature and moisture control CO~2~ and CH~4~ emissions from mineral soils sampled from the bottom of the annual active layer, i.e. directly above permafrost, in an Alaskan boreal forest. Gas emissions from thirty cores, subjected to two temperatures and either field moisture conditions or experimental drought, were tracked over a 100-day incubation; we also measured a variety of physical and chemical characteristics of the cores. Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (unitless) at the beginning of the incubation; cores at field moisture were unchanged at the end, but drought cores had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Carbon dioxide fluxes were strongly influenced by incubation chamber temperature, core water content, and percent soil nitrogen, and had a temperature sensitivity (i.e. Q~10~) of `r pn(Q10_cumulative["Field moisture", "CO2"], 1)` and `r pn(Q10_cumulative["Drought", "CO2"], 1)` for the field moisture and drought treatments, respectively.. Methane fluxes were most strongly correlated with percent nitrogen, but neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes. The cumulative production of C from CO~2~ was over `r minmag` orders of magnitudes higher than that from CH~4~. These results suggest that deep active-layer soils may be much more sensitive to changes in moisture than to temperature, a critical factor as discontinuous permafrost melts in interior Alaska. Deep but unfrozen high-latitude soils have been shown to be strongly affected by long-term experimental warming, and these results provide insight into their future dynamics and feedback potential with future climate change.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented increased fluxes of CO~2~, and under some conditions of CH~4~ {Olefeldt, 2013 #3456}, with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. Greenhouse gas (GHG) responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. Decadal warming and drying trends in Alaska {Bieniek, 2014 #4389} may counteract these effects, however. A critical question, then, is how the structure, chemistry, and microbial communities of current active-layer soils will respond to almost-certain temperature and moisture changes in the future {Xue, 2016 #4361}.

@@

The field component of this research took place in Caribou-Poker Creeks Research Watershed (CPCRW), part of the Bonanza Creek LTER (http://www.lter.uaf.edu/research/study-sites-cpcrw). CPCRW is located in the Yukon-Tanana Uplands northeast of Fairbanks, AK, a part of the boreal forest that has seen strong increases in air temperature and forest browning {Ju, 2016 #4320} over several decades. Annual average air temperature is -2.5 °C, and annual average precipitation 400 mm {Petrone, 2006 #4313}. The watershed's lowlands and north-facing slopes are dominated by black spruce (*Picea mariana* (Mill.) BSP), feathermoss (*Pleurozium schreberi* and others), and *Sphagnum* spp.; the drier south slopes tend to be deciduous with a mixture of trembling aspen (*Populus tremuloides* Michx.), paper birch (*Betula neoalaskana*), and patches of alder (*Alnus crispa*).

@@

Thirty-nine soil cores, each 30 cm high by 7.5 cm wide, were taken using a soil recovery augur (AMS Inc., American Falls, ID) on 3-5 August 2015. We sampled from the bottom (within 0-2 cm of permafrost) of the active layer, which averaged 80 cm depth. Sample points were randomly located in the transition zone described above, and separated by 2-5 m. Cores were kept cool in the field before being packed in dry ice and shipped to Richland, WA within 48-72 hours of collection.

@@

On 18 August 2015 cores were placed into one of two growth chambers (Conviron Control Systems BDW80, Winnipeg, Canada) maintained at 4 and 20 °C temperatures and 70% relative humidity and allowed to equilibrate for two weeks. Starting on 31 August 2015 we measured the cores' mass and GHG (CH~4~ and CO~2~) emissions four times in the first week, then twice per week for the first month, and then once per week for the rest of the 100-day incubation. Throughout the incubation, cores had a 200 µm mesh screen fit to the base and were mounted on porous ceramic plates (Soil Moisture Equipment Corp., Santa Barbara, CA, USA) so that, when the plates were placed in contact with water, water would move up into the cores via capillary action. The "drought" cores were mounted on dry plates, but not allowed to drop below 5% water content. When necessary, cores received additional wetting from the top to maintain their water status at the desired level.

@@

@@

For each measurement of each sample throughout the 100-day incubation (i.e., each gas, core, and date/time), we used the rise in gas concentrations to calculate a flux rate in ppm s^-1^ (CO~2~) or ppb s^-1^ (CH~4~), a linear rate of change (δc/δt) based on the concentration rise from a minimum (up to `r MAX_MINCONC_TIME` seconds after measurement began) to a maximum (at `r MAX_MINCONC_TIME`-`r MAX_MAXCONC_TIME` seconds). Each core’s respiration flux (*F*) was then calculated as $F=\frac{\delta c}{\delta t} \frac{V}{M} \frac{P_a}{RT}$ where *V* is the core-specific system volume, *M* the core dry mass as determined at the end of the incubation, *P~a~* atmospheric pressure (101 kPa; the incubation chambers were ~120 m a.s.l.), *R* the universal gas constant (8.3 x 10^-3^ m^3^ kPa mol^-1^ K^-1^) and *T* the chamber air temperature (K) at time of measurement. The final respiration rate was expressed on a dry soil mass basis (µg C g soil^-1^ day^-1^).

@@

Cumulative respiration for each core and gas was calculated by linearly interpolating flux rates between measurement dates and summing respired C over the entire incubation. The effect of temperature and treatment (drought, controlled drought, or field moisture conditions) on cumulative gas fluxes was evaluated with a post-hoc Tukey Honest Significant Differences test. Temperature sensitivity (Q~10~) was calculated for each gas and treatment as $\frac{F_2}{F_1}^{\left( \frac{10}{T_2-T_1} \right)}$ where *F~1~* and *F~2~* are the cumulative gas fluxes (mg C g C^-1^) at temperatures *T~1~* and *T~2~* (°C), respectively.

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" {Bond-Lamberty, 2016 #4388} with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [*DOI to be filled in on acceptance*].

@@

Gravimetric water content was `r pn(gwcbegin$mean, 2)` ± `r pn(gwcbegin$sd, 2)` (min `r pn(gwcbegin$min, 2)`, max `r pn(gwcbegin$max, 2)`) at the beginning of the incubation (**Figure 1**). "Field moisture" cores were on average unchanged (`r pn(gwcendmean["Field moisture"], 2)` ± `r pn(gwcendsd["Field moisture"], 2)`) at the end of the incubation, but both the drought treatments, which did not differ from each other in their effect on gravimetric water content (P = `r pn(p_trt_wc, 3)`), had declined to `r pn(gwcendmean["Drought"], 2)` ±  `r pn(gwcendsd["Drought"], 2)`. Volumetric water content values ranged from `r pn(vwcbegin$mean, 2)` ± `r pn(vwcbegin$sd, 2)` (min `r pn(vwcbegin$min, 2)`, max `r pn(vwcbegin$max, 2)`) at the beginning of the experiment to `r pn(vwcend$mean, 2)` ± `r pn(vwcend$sd, 2)` (min `r pn(vwcend$min, 2)`, max `r pn(vwcend$max, 2)`) at the end across all cores. Water filled pore space, assuming a particle density of `r PARTICLE_DENSITY` g cm^-3^, was `r pn(min(fluxdata$WFPS_percent), 0)`-`r pn(max(fluxdata$WFPS_percent), 0)`% over all cores, moisture treatments, and temperatures.

@@

Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes.

@@

We found that CO~2~, but not CH~4~, fluxes from these oxic active-layer mineral soils were highly sensitive to temperature and drought effects. Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both temperature and moisture, consistent with our results They found however that CH~4~ uptake (no emissions were observed) was driven by temperature only. Waldrop et al. {, 2010 #4380} incubated active-layer and permafrost soils from *Picea mariana* sites near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost) from -5 to 5 °C; these values are higher than we observed, consistent with the lower temperature range {Hamdi, 2013 #3664} and fundamental biokinetics {Davidson, 2006 #1669}. Waldrop et al. {, 2010 #4380} also observed flux rates of 0.001-0.10 µmol CH~4~ day^-1^ g^-1^ (~`r pn(0.001 * 16 / 0.012 / 1000, 3)`-`r pn(0.1 * 16 / 0.012 / 1000, 3)` ng C g C^-1^ day^-1^), differing by orders of magnitude between sites (but roughly similar to our observed CH~4~ emissions), and ~1-5 µg C-CO2 hr^-1^ g^-1^ (~`r pn(1 * 24 / 0.012, 0)`-`r pn(5 * 24 / 0.012, 0)` µg C g C^-1^ day^-1^), considerably higher than the CO~2~ rates observed from our cores. In an incubation of active-layer Alaskan permafrost peats, Treat et al. {, 2014 #4213} found CO~2~ and CH~4~ emissions to be strongly correlated with temperature and moisture. Finally, during the first 100 days of a year-long incubation of Fairbanks-area 0-10 cm mineral soils, Neff and Hooper {, 2002 #4390} observed fluxes of ~`r pn(0.2 * 1000 * 12/44, 0)`-`r pn(1.5 * 1000 * 12/44, 0)` µg C-CO~2~ g C^-1^ day^-1^, in line with the results here.
